---
title: <font size="7"><b>Data analysis</b></font>
subtitle: <font size="4"><b>Rensch's Rule in Costa Rican hummingbirds</b><br><br>University of Costa Rica</font>
author: <font size="3"><a href="http://marceloarayasalas.weebly.com/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    css: extra.css
    df_print: tibble
    highlight: pygments  
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
fontsize: 12pt 
editor_options: 
  chunk_output_type: console
---

<!-- this code add line numbers to code blocks -->
<!-- only works when code folding is not used in yaml (code_folding: show) -->

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

```{r load packages and setup style, echo = FALSE, message = FALSE, warning=FALSE}

# github packages must include user name ("user/package")
# bioconductor packages must include "bioc" ("bioc/package")
# knitr is require for creating html/pdf/word reports
# kableExtra is used to print pretty formatted tables 
# formatR is used for soft-wrapping code
# xaringanExtra::use_clipboard is used for adding a copy button to each code block
pkgs <- c("kableExtra", "knitr", "formatR", "rprojroot", "xaringanExtra", "ape", "readxl", bioconductor = "ggtree", github = "maRce10/brmsish", github = "maRce10/sketchy", "brms", "viridis")

# install/ load packages
sketchy::load_packages(pkgs, quite = TRUE)

# set working directory as project directory or one directory above,
rootdir <- try(rprojroot::find_rstudio_root_file(), silent = TRUE)
if (is(rootdir, "try-error")) rootdir <-  ".."
opts_knit$set(root.dir = rootdir)

# options to customize chunk outputs
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", # for code line numbers
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE
 )

# this is a customized printing style data frames 
# screws up tibble function
tibble <- function(x, ...) { 
  x <- kbl(x, digits=4, align= 'c', row.names = FALSE) 
   x <- kable_styling(x, position ="center", full_width = FALSE,  bootstrap_options = c("striped", "hover", "condensed", "responsive")) 
   asis_output(x)
}

registerS3method("knit_print", "data.frame", tibble)

# to add copy button to code blocks
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)

source("~/Dropbox/R_package_testing/brmsish/R/html_summary.R")
```

&nbsp; 

<!-- skyblue box -->
<div class="alert alert-info">

# Purpose

- Explore sexual size dimorphism in hummingbird species from Costa Rica

- Evaluate Rensch's Rule

</div>

&nbsp; 


<!-- light brown box -->
<div class="alert alert-warning">

# Report overview

- You can have the sections listed here, for instance:

  - [Explore data](#explore-data)
  - [Run statistical models](#run-statistical-models)
  - [Size comparison among sexes](#size-comparison-among-sexes)
  - [SSD (Lovich-Gibbons ratio) vs male size](#ssd-(lovich-gibbons ratio)-vs-male-size)

</div>

&nbsp;

# Explore data

```{r, eval = TRUE}

# read data
dat <- as.data.frame(read_excel("./data/raw/supplementary data.xlsx"))

# read tree
tree <- read.tree("./data/raw/imputed hummer tree 339 spp.tre")

dat$`scientific name`[dat$`scientific name` == "Phaetornis guy"] <- "Phaethornis guy"
dat$`scientific name`[dat$`scientific name` == "Phaetornis longirostris"] <- "Phaethornis longirostris"
dat$`scientific name`[dat$`scientific name` == "Heliomaster longisrostris"] <- "Heliomaster longirostris"
dat$scientific_name <- gsub(" ", "_", dat$`scientific name`)

tree$tip.label[tree$tip.label == "Amazilia_amabilis"] <- "Polyerata_amabilis"
tree$tip.label[tree$tip.label == "Hylocharis_eliciae"] <- "Chlorestes_eliciae"
tree$tip.label[tree$tip.label == "Chlorostilbon_canivetii"] <- "Cynanthus_canivetii"
tree$tip.label[tree$tip.label == "Elvira_cupreiceps"] <- "Microchera_cupreiceps"
tree$tip.label[tree$tip.label == "Calliphlox_bryantae"] <- "Philodice_bryantae"
tree$tip.label[tree$tip.label == "Amazilia_edward"] <- "Saucerottia_edward"
tree$tip.label[tree$tip.label == "Amazilia_saucerottei"] <- "Saucerottia_hoffmanni"
tree$tip.label[tree$tip.label == "Amazilia_saucerottei"] <- "Saucerottia_hoffmanni"
tree$tip.label[tree$tip.label == "Amazilia_candida"] <- "Chlorestes_candida"
tree$tip.label[tree$tip.label == "Elvira_chionura"] <- "Microchera_chonura"

# setdiff(unique(dat$scientific_name), tree$tip.label)

subtree <- drop.tip(phy = tree, tip = setdiff(tree$tip.label, unique(dat$scientific_name))
)

names(dat) <- gsub(" ", ".", names(dat))

subtree2 <- subtree
subtree2$tip.label <- gsub("_", " ", subtree2$tip.label)
ggtree(subtree2, layout = "circular", col = viridis(10)[7]) + geom_tiplab(size = 4) + theme(plot.margin = unit(c(30, 10, 30, 10), "mm"))

```

# Run statistical models

## Size comparison among sexes
### log(females) ~ log(males) 

(note that in the data supplied ln.SEX.weight variables are in natural log)

```{r, eval = TRUE}

# model settings
v.cv.tree <- ape::vcv.phylo(subtree)

prior <- c(
    prior(normal(0, 10), "b"),
    prior(normal(0, 50), "Intercept"),
    prior(student_t(3, 0, 20), "sd"),
    prior(student_t(3, 0, 20), "sigma")
  )

```

```{r, eval = FALSE}

male_x_mod <- brm(
  ln.female.weight ~ ln.male.weight  + (1|gr(scientific_name, cov = v.cv.tree)),
  data = dat,
  family = gaussian(),
  data2 = list(v.cv.tree = v.cv.tree),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  file =  "./data/processed/male_vs_female_size"
)

```

```{r, message=FALSE, warning=FALSE, results='asis'}

html_summary(read.file = "./data/processed/male_vs_female_size.rds", n.posterior = 1000)

```

### log(males) ~ log(females) 

```{r, eval = FALSE}

female_x_mod <- brm(
  ln.male.weight ~ ln.female.weight  + (1|gr(scientific_name, cov = v.cv.tree)),
  data = dat,
  family = gaussian(),
  data2 = list(v.cv.tree = v.cv.tree),
   prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  file =  "./data/processed/female_vs_male_size.RDS"
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

html_summary(read.file = "./data/processed/female_vs_male_size.RDS", n.posterior = 1000)

```

## SSD (Lovich-Gibbons ratio) vs male size

- SSD represented as aboslute Lovich-Gibbons ratio
- sex size as log of weight

### Absolute SSD vs log male size
```{r, eval = FALSE}

male_vs_ssd_mod <- brm(
  aboluteSSD ~ ln.male.weight  + (1|gr(scientific_name, cov = v.cv.tree)),
  data = dat,
  family = gaussian(),
  data2 = list(v.cv.tree = v.cv.tree),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  s =  "./data/processed/male_vs_abs_SSD.RDS")

```

```{r, message=FALSE, warning=FALSE, results='asis'}

html_summary(read.file = "./data/processed/male_vs_abs_SSD.RDS", n.posterior = 1000)

```

### absolute SSD vs log female size
```{r, eval = FALSE}

female_vs_ssd_mod <- brm(
  aboluteSSD ~ ln.female.weight  + (1|gr(scientific_name, cov = v.cv.tree)),
  data = dat,
  family = gaussian(),
  data2 = list(v.cv.tree = v.cv.tree),
   prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  file = "./data/processed/female_vs_abs_SSD.RDS")

```

```{r, message=FALSE, warning=FALSE, results='asis'}

html_summary(read.file = "./data/processed/female_vs_abs_SSD.RDS", n.posterior = 1000)

```

<!-- '&nbsp;' adds blank space -->



<!-- light green box -->

<div class="alert alert-success">

&nbsp; 

# Takeaways

 - Dimorpism increases with body size in males but not in females



</div>

&nbsp;

<!-- '---' adds a gray vertical line -->

---

&nbsp; 
 
 <!-- add packages used, system details and versions  -->
 
<font size="4">Session information</font>

```{r session info, echo=F}

sessionInfo()

```
