---
title: <font size="7"><b>Data analysis</b></font>
subtitle: <font size="4"><b>Rensch's Rule in Costa Rican hummingbirds</b><br>University of Costa Rica</font>
author: <font size="3"><a href="https://marce10.github.io/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
toc: true
toc-depth: 2
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: show
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>
```

```{r set root directory, echo = FALSE}

# set working directory as project directory or one directory above,
rootdir <- try(rprojroot::find_rstudio_root_file(), silent = TRUE)
if (is(rootdir, "try-error")) rootdir <-  ".."
knitr::opts_knit$set(root.dir = rootdir)

```

```{r add link to github repo, echo = FALSE, results='asis'}

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  cat("\nSource code, data and version control of the project found  at [", url, "](", url, ")", sep = "")
  }

```

```{r load packages and setup style, echo = FALSE, message = FALSE, warning=FALSE}

# github packages must include user name ("user/package")
# bioconductor packages must include "bioc" ("bioc/package")
# knitr is require for creating html/pdf/word reports
# kableExtra is used to print pretty formatted tables 
# formatR is used for soft-wrapping code
# xaringanExtra::use_clipboard is used for adding a copy button to each code block
pkgs <- c("kableExtra", "knitr", "formatR", "rprojroot", "xaringanExtra", "ape", "readxl", bioconductor = "ggtree", github = "maRce10/brmsish", github = "maRce10/sketchy", "brms", "viridis", "ggplot2", "phangorn", "gghalves")

# install/ load packages
sketchy::load_packages(pkgs, quite = TRUE)

# set working directory as project directory or one directory above,
rootdir <- try(rprojroot::find_rstudio_root_file(), silent = TRUE)
if (is(rootdir, "try-error")) rootdir <-  ".."
opts_knit$set(root.dir = rootdir)

# options to customize chunk outputs
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", # for code line numbers
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE,
  warning = FALSE
 )

# to add copy button to code blocks
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)


```

&nbsp; 

<!-- skyblue box -->
<div class="alert alert-info">

# Purpose {.unnumbered .unlisted}

- Explore sexual size dimorphism in hummingbird species from Costa Rica

- Evaluate Rensch's Rule

</div>

&nbsp; 


<!-- light brown box -->
<div class="alert alert-warning">

# Report overview {.unnumbered .unlisted}

- You can have the sections listed here, for instance:

  - [Explore data](#explore-data)
  - [Run statistical models](#run-statistical-models)
  - [Size comparison among sexes](#size-comparison-among-sexes)
  - [SSD (Lovich-Gibbons ratio) vs male size](#ssd-(lovich-gibbons ratio)-vs-male-size)

</div>

&nbsp;

# Explore data

```{r, eval = TRUE}

# read data
dat <- as.data.frame(read_excel("./data/raw/supplementary data.xlsx"))

dat$Clade[dat$Clade == "Billiants"] <- "Brilliants"

# read trees
trees <- read.tree("./data/raw/Posterior trees fixed names 339 species 1 swift.trees")

dat$`scientific name`[dat$`scientific name` == "Phaetornis guy"] <- "Phaethornis guy"
dat$`scientific name`[dat$`scientific name` == "Phaetornis longirostris"] <- "Phaethornis longirostris"
dat$`scientific name`[dat$`scientific name` == "Heliomaster longisrostris"] <- "Heliomaster longirostris"
dat$scientific_name <- gsub(" ", "_", dat$`scientific name`)

trees <- lapply(trees, function(tree){
    tree$tip.label[tree$tip.label == "Amazilia_amabilis"] <- "Polyerata_amabilis"
    tree$tip.label[tree$tip.label == "Hylocharis_eliciae"] <- "Chlorestes_eliciae"
    tree$tip.label[tree$tip.label == "Chlorostilbon_canivetii"] <- "Cynanthus_canivetii"
    tree$tip.label[tree$tip.label == "Elvira_cupreiceps"] <- "Microchera_cupreiceps"
    tree$tip.label[tree$tip.label == "Calliphlox_bryantae"] <- "Philodice_bryantae"
    tree$tip.label[tree$tip.label == "Amazilia_edward"] <- "Saucerottia_edward"
    tree$tip.label[tree$tip.label == "Amazilia_saucerottei"] <- "Saucerottia_hoffmanni"
    tree$tip.label[tree$tip.label == "Amazilia_saucerottei"] <- "Saucerottia_hoffmanni"
    tree$tip.label[tree$tip.label == "Amazilia_candida"] <- "Chlorestes_candida"
    tree$tip.label[tree$tip.label == "Elvira_chionura"] <- "Microchera_chonura"
    
    subtree <- drop.tip(phy = tree, tip = setdiff(tree$tip.label, unique(dat$scientific_name))
)

    return(tree)
})


names(dat) <- gsub(" ", ".", names(dat))

# maximum clade credibility consensus tree
mcd_tree <- phangorn::maxCladeCred(trees)

mcd_tree$tip.label <- gsub("_", " ", mcd_tree$tip.label)

ggtree(mcd_tree, layout = "circular", col = viridis(10)[7]) + geom_tiplab(size = 2) + theme(plot.margin = unit(c(30, 10, 30, 10), "mm"))

# add log 10 variables
dat$lg10.female.weight <- log10(dat$female.weight)
dat$lg10.male.weight <- log10(dat$male.weight)

```

# Run statistical models

```{r, eval = TRUE}

prior <- c(
    prior(normal(0, 10), "b"),
    prior(normal(0, 50), "Intercept"),
    prior(student_t(3, 0, 20), "sd"),
    prior(student_t(3, 0, 20), "sigma")
  )

iter <- 10000

chains <- 1

cores <- 1
```

## Size comparison among sexes
### log(females) ~ log(males) 

(note that in the data supplied ln.SEX.weight variables are in natural log)

```{r, eval = FALSE}

male_x_mod <- phylogenetic_uncertainty(
  ln.female.weight ~ ln.male.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "male_vs_female_size",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
)

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/male_vs_female_size_1000_iterations.rds", highlight = TRUE)

```


### log10(females) ~ log10(males) 

```{r, eval = FALSE}

log10_male_x_mod <- phylogenetic_uncertainty(
  lg10.female.weight ~ lg10.male.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "log10_male_vs_female_size",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
)

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/log10_male_vs_female_size_1000_iterations.rds", highlight = TRUE)

```


### log(males) ~ log(females) 

```{r, eval = FALSE}

phylogenetic_uncertainty(
  ln.male.weight ~ ln.female.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "female_vs_male_size",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/female_vs_male_size_1000_iterations.rds", highlight = TRUE)

```

### log10(males) ~ log10(females) 

```{r, eval = FALSE}

phylogenetic_uncertainty(
  lg10.male.weight ~ lg10.female.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "log10_female_vs_male_size",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/log10_female_vs_male_size_1000_iterations.rds", highlight = TRUE)

```


## SSD (Lovich-Gibbons ratio) vs male size

- SSD represented as aboslute Lovich-Gibbons ratio
- sex size as log of weight

### Absolute SSD vs log male size
```{r, eval = FALSE}

phylogenetic_uncertainty(
  aboluteSSD ~ ln.male.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "male_vs_abs_SSD",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
   )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/male_vs_abs_SSD_1000_iterations.rds", highlight = TRUE)

```

### Absolute SSD vs log10 male size
```{r, eval = FALSE}

phylogenetic_uncertainty(
  aboluteSSD ~ lg10.male.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "log10_male_vs_abs_SSD",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/log10_male_vs_abs_SSD_1000_iterations.rds", highlight = TRUE)

```

### absolute SSD vs log female size
```{r, eval = FALSE}

phylogenetic_uncertainty(
  aboluteSSD ~ ln.female.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "female_vs_abs_SSD",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/female_vs_abs_SSD_1000_iterations.rds", highlight = TRUE)

```

### absolute SSD vs log10 female size
```{r, eval = FALSE}

phylogenetic_uncertainty(
  aboluteSSD ~ lg10.female.weight,
  data = dat,
  sp.id.column = "scientific_name",
  phylos = trees,
  family = gaussian(),
  prior = prior,
  iter = iter,
  control=list(adapt_delta=0.99, max_treedepth=15),
  fit.name = "log10_female_vs_abs_SSD",  
  save.fits = FALSE,
  save.combined = TRUE,
  path = "./data/processed/", 
  chains = chains, 
  cores = cores
  )

```

```{r, message=FALSE, warning=FALSE, results='asis'}

brmsish::extended_summary(read.file = "./data/processed/log10_female_vs_abs_SSD_1000_iterations.rds", highlight = TRUE)

```


# Graphs

## Male vs female body mass

```{r}

fit <- readRDS("./data/processed/log10_male_vs_female_size_1000_iterations.rds")

gg_fit <- conditional_effects(fit)

pred_dat <- gg_fit$lg10.male.weight 
pred_dat$Clade <- "Hermits"

fit_summ <- brmsish:::draw_summary(draws = posterior::as_draws_array(fit), variables = c("b_Intercept",  "b_lg10.male.weight"), robust = TRUE, probs = c(0.025, 0.975), spread.type = "MAD")


ggplot(data = dat, mapping = aes(x = lg10.male.weight, y = lg10.female.weight, fill = Clade, shape = Clade)) +
    scale_fill_viridis_d(alpha = 0.5) +
    geom_segment(data = pred_dat, mapping = aes(x = min(pred_dat$lg10.male.weight), y = min(pred_dat$estimate__), xend = max(pred_dat$lg10.male.weight), yend = max(pred_dat$estimate__))) +
    geom_ribbon(data = pred_dat, aes(ymin = lower__, ymax =  upper__), fill = "gray", alpha = 0.3) +
    scale_shape_manual(values = c(21, 22, 23, 21, 22, 23, 21, 22)) +
    geom_point(size = 3, color = "transparent") +
    theme_classic(base_size = 20) + labs(x = "log10(male body mass)", y = "log10(female body mass)")

# ggsave(filename = "./output/log10_male_vs_absolute_SSD_70dpi.tiff", dpi = 70)
# 
# ggsave(filename = "./output/log10_male_vs_absolute_SSD_300dpi.tiff", dpi = 300)


```


## Male body mass vs absolute SSD

```{r}

fit <- readRDS("./data/processed/log10_male_vs_abs_SSD_1000_iterations.rds")

gg_fit <- conditional_effects(fit)

pred_dat <- gg_fit$lg10.male.weight
pred_dat$Clade <- "Hermits"

fit_summ <- brmsish:::draw_summary(draws = posterior::as_draws_array(fit), variables = c("b_Intercept",  "b_lg10.male.weight"), robust = TRUE, probs = c(0.025, 0.975), spread.type = "MAD")

ggplot(data = dat, mapping = aes(x = lg10.male.weight, y = aboluteSSD, fill = Clade, shape = Clade)) +
    scale_fill_viridis_d(alpha = 0.5) +
    geom_segment(data = pred_dat, mapping = aes(x = min(pred_dat$lg10.male.weight), y = min(pred_dat$estimate__), xend = max(pred_dat$lg10.male.weight), yend = max(pred_dat$estimate__))) +
    geom_ribbon(data = pred_dat, aes(ymin = lower__, ymax =  upper__), fill = "gray", alpha = 0.3) +
    scale_shape_manual(values = c(21, 22, 23, 21, 22, 23, 21, 22)) +
    geom_point(color = viridis(10, alpha = 0.6)[7], size = 3) +
    theme_classic(base_size = 20) + labs(x = "log10(male body mass)", y = "Dimorphism (absolute SSD)")

# ggsave(filename = "./output/log10_female_vs_absolute_SSD_300dpi.tiff", dpi = 300)
# 
# ggsave(filename = "./output/log10_female_vs_absolute_SSD_70dpi.tiff", dpi = 70)

```

## Female body mass vs absolute SSD

```{r}

fit <- readRDS("./data/processed/log10_female_vs_abs_SSD_1000_iterations.rds")

gg_fit <- conditional_effects(fit)

pred_dat <- gg_fit$lg10.female.weight
pred_dat$Clade <- "Hermits"         

fit_summ <- brmsish:::draw_summary(draws = posterior::as_draws_array(fit), variables = c("b_Intercept",  "b_lg10.female.weight"), robust = TRUE, probs = c(0.025, 0.975), spread.type = "MAD")

ggplot(data = dat, mapping = aes(x = lg10.female.weight, y = aboluteSSD, fill = Clade, shape = Clade)) +
    scale_fill_viridis_d(alpha = 0.5) +
    # geom_segment(data = pred_dat, mapping = aes(x = min(pred_dat$lg10.female.weight), y = min(pred_dat$estimate__), xend = max(pred_dat$lg10.female.weight), yend = max(pred_dat$estimate__))) +
    # geom_ribbon(data = pred_dat, aes(ymin = lower__, ymax =  upper__), fill = "gray", alpha = 0.3) +
    scale_shape_manual(values = c(21, 22, 23, 21, 22, 23, 21, 22)) +
    geom_point(color = viridis(10, alpha = 0.6)[7], size = 3) +
    theme_classic(base_size = 20) + labs(x = "log10(female body mass)", y = "Dimorphism (absolute SSD)")

# ggsave(filename = "./output/male_vs_female_size_10log_300dpi.tiff", dpi = 300)
# ggsave(filename = "./output/male_vs_female_size_10log_70dpi.tiff", dpi = 70)

```

```{r}

fill_color <- viridis(10)[7]

dat$Clade <- factor(dat$Clade, levels = c("Bees", "Coquettes", "Hermits", "Brilliants", "Emeralds", "Mangoes", "Mountain Gems", "Topazes"))

agg_dat <- aggregate(`scientific_name` ~ Clade, data = dat, length)
agg_dat$`Lovich-Gibbons.ratio` <- NA
agg_dat$n.labels <- paste("n =", agg_dat$`scientific_name`)

# set.seed(12)
trans <-  function(height = 0, ...) ggplot2::position_jitter(height = height, ...)

# composed box plot
ggplot(dat, aes(y = `Lovich-Gibbons.ratio`, x = Clade)) +
    geom_hline(yintercept = 0, lty = 2) +     
## add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    fill = fill_color,
    alpha = 0.5,
    ## custom bandwidth
    adjust = .5,
    ## adjust height
    width = .6,
    .width = 0,
    ## move geom to the cright
    justification = -.2,
    point_colour = NA
  ) +
  geom_boxplot(fill = fill_color,
    width = .15,
    ## remove outliers
    outlier.shape = NA ## `outlier.shape = NA` works as well
  ) +
  ## add justified jitter from the {gghalves} package
  # geom_half_point(
  #   color = fill_color,
  #   ## draw jitter on the left
  #   side = "l",
  #   ## control range of jitter
  #   range_scale = .4,
  #   ## add some transparency
  #   alpha = .5,
  #   transformation = ggplot2::position_jitter(height = 0)
  # 
  # ) +
    labs(x="Clade", y= "Lovich-Gibbons ratio"
  ) +
  ylim(c(-0.39, 0.145)) +
  geom_text(data = agg_dat, aes(y = rep(-0.387, nrow(agg_dat)), x = Clade, label = n.labels), nudge_x = 0, size = 6) +
     theme_classic(base_size = 18) +
theme(axis.text.x = element_text(angle = 30, hjust = 1)) 

```


# Combined model diagnostics

```{r, results='asis'}

check_rds_fits(path = "./data/processed", html = TRUE)

```



<!-- light green box -->

<div class="alert alert-success">

# Takeaways {.unnumbered .unlisted}

 - Dimorpism increases with body size in males but not in females

</div>

&nbsp;

<!-- '---' adds a gray vertical line -->

---

&nbsp; 
 
 <!-- add packages used, system details and versions  -->
 
# Session information {.unnumbered .unlisted}

```{r session info, echo=F}

sessionInfo()

```
    
